Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/relatedwork.tex
more in related work

In the partial evaluation community, stage splitting has been investigated 
mostly as a preprocessing step for self-applied partial evaluators. 
For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
for programs consistening of a set of first-order recursive functions.
Those works used a grammar-based binding-time specification that's 
relatively complicated compared to the type system of \lang.
This complication mostly stems from the decision to start with an untyped language;
.

In addition to the lack of first-class functions, 
the binding time specificiations used by these systems are more complicated and less expressive than the type system of \lang.

The question of binding time analysis for a higher-order language was visited in \cite{Mogensen89b},
however the implications for splitting were not explicitly considered.

Automatic pass separation has also been a topic of interest in
programming systems for computer graphics, where to achieve
high performance, programs are typically manually separated into
components by frequency of execution. (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  
The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
suggestions of differing the task of splitting to the compiler.
Early work (\cite{knoblock96}) sought to determine the staging of programs automatically through use of binding time analysis,
however later work (\cite{Proudfoot:2001,Foley:2011,He:2014}) has focused on giving programmers explicit control of the staging through 
programming language features. 
All of these efforts in graphics have been limited to simple, imperative languages.

>>>>

In the partial evaluation community, stage splitting has been investigated 
mostly as a preprocessing step for self-applied partial evaluators. 
For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
for programs consistening of a set of first-order recursive functions.
Those works use a grammar-based binding-time specification that's 
relatively complicated compared to the type system of \lang.
This complication mostly arises from the decision to start with an untyped language:
one must first find a way to say ``argument $x$ will be a tuple'' before one can say
``the first component of $x$ is available at the second stage,'' 
and so much of their binding-time specification just amounts to a type system.
Formulations more like \lang\ (namely the two-level system of \cite{NN92-twolevel}) were known at the time,
but \cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
Davies essentially solved this limitation in \cite{davies96} (pending our further tweak with $\curr$),
and the question of stage splitting as we have presented it has been ripe ever since.
Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b},
but the implications for splitting were not explicitly considered.

Splitting-like algorithms have also been a topic of interest in
programming systems for computer graphics, where to achieve
high performance, programs are typically manually separated into
components by frequency of execution. (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  
The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
suggestions of defering the task of splitting to the compiler.
Early work \cite{knoblock96} sought to determine the staging of programs automatically through use of binding time analysis,
however later work \cite{Proudfoot:2001,Foley:2011,He:2014} has focused on giving programmers explicit control of the staging through 
programming language features. 
All of these efforts in graphics have been limited to simple, imperative languages.
